Importance of depth refugia for reef resilience and intervention on the Great Barrier Reef under future climate change

### 2025-09-25 paper/paper_draft.qmd
Update working title, author list and acknowledgements section

---
title: Think deep for reef resilience
author:
  - name: Benjamin Grier
    email: grierben777@gmail.com
    affiliations:
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Cape Cleveland Rd
          state: QLD
    attributes:
        corresponding: true
  - name: Takuya Iwanaga
    email: t.iwanaga@aims.gov.au
    affiliations:
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Cape Cleveland Rd
          city: Cape Cleveland
          state: QLD
    attributes:
        corresponding: true
abstract: |
    Coral reefs globally face unprecedented threats from climate change, with the Great Barrier
    Reef (GBR) experiencing cumulative stressors and increasingly severe declines in coral
    cover from thermal stress events.
    Understanding drivers behind reef resilience to climate impacts is critical for conservation
    planning and intervention strategies. A scenario-based population modelling approach was
    adopted with larval connectivity dynamics, thermal adaptation, and environmental factors
    to assess coral reef resilience across GBR locations under future climate projections.
    Using five Global Climate Models (GCMs) spanning five Equilibrium Climate Sensitivities,
    we analysed reef coral cover time series and the ability of reefs to maintain positive carbonate
    production budgets under future conditions. Contrary to expectations, larval connectivity
    patterns did not correlate with increases in maintenance of positive carbonate budgets.
    Instead, reef depth emerged as the primary predictor of cover-time series resilience and
    maintenance of positive carbonate budgets, with deeper reefs benefiting from thermal
    refugia that reduce bleaching impacts. Interaction between reef depth and incoming
    connectivity was found to influence reef temporal behaviour at 15m depth. Our findings suggest that
    depth provides critical protection against thermal stress, with high incoming connectivity enhancing the
    performance of deep reefs, contrasting current intervention
    practices that primarily target easily accessible sites in the 1-10m depth range. These results have
    important implications for reef management, indicating that depth refugia should be
    prioritized for further consideration in conservation planning alongside connectivity
    considerations to maximize coral survival under continuing climate change.
keywords:
  - Coral reef
  - Climate change
  - Ecosystem resilience
  - Reef depth
  - Reef connectivity
date: last-modified
format:
  elsevier-docx:
    journal:
      name: Journal Name
      formatting: preprint
      cite-style: authoryear
bibliography: references.bib
engine: julia
crossref:
  custom:
  - kind: float
    reference-prefix: SFigure
    key: sfig
  - kind: float
    reference-prefix: STable
    key: stbl
editor:
  markdown:
    wrap: 92
    canonical: true
---

@@

**Benjamin Grier:** conceptualisation, data curation, formal analysis, investigation,
methodology, visualization, writing – original draft, writing – review and editing. **Takuya
Iwanaga:** conceptualisation, data curation, formal analysis, investigation, methodology,
visualization, writing – original draft, writing – review and editing. **\[BG: To be
completed.\]**

@@

We acknowledge the contributions of Barbara Robson and Jessica Benthuysen to manuscript
preparation. This research was funded by the Reef Restoration and Adaptation Program, and
facilitated by the Australian Institute of Marine Science. We acknowledge thermal projection
data sourced from HighResCoralStress datasets, model parameterisation data sourced from the
Ecological Intelligence for Reef Restoration (EcoRRAP) subprogram, larval transition
probability data derived from the eReefs GBR1 hydrodynamic model, and spatial and depth data
available from the Great Barrier Reef Marine Park Authority.

>>>>

---
title: Importance of depth refugia for reef resilience and intervention on the Great Barrier Reef under future climate change
author:
  - name: Benjamin Grier
    email: grierben777@gmail.com
    affiliations:
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Cape Cleveland Rd
          state: QLD
    attributes:
        corresponding: true
  - name: Takuya Iwanaga
    email: t.iwanaga@aims.gov.au
    affiliations:
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Cape Cleveland Rd
          city: Cape Cleveland
          state: QLD
  - name: Chinenye Ani
    email: c.ani@aims.gov.au
    affiliations:
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Cape Cleveland Rd
          city: Cape Cleveland
          state: QLD
  - name: Pedro Ribeiro De Almeida
    email: p.ribeirodealmeida@aims.gov.au
    affiliations:
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Cape Cleveland Rd
          city: Cape Cleveland
          state: QLD
  - name: Sam Matthews
    email: s.matthews@aims.gov.au
    affiliations:
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Cape Cleveland Rd
          city: Cape Cleveland
          state: QLD
abstract: |
    Coral reefs globally face unprecedented threats from climate change, with the Great Barrier
    Reef (GBR) experiencing cumulative stressors and increasingly severe declines in coral
    cover from thermal stress events.
    Understanding drivers behind reef resilience to climate impacts is critical for conservation
    planning and intervention strategies. A scenario-based population modelling approach was
    adopted with larval connectivity dynamics, thermal adaptation, and environmental factors
    to assess coral reef resilience across GBR locations under future climate projections.
    Using five Global Climate Models (GCMs) spanning five Equilibrium Climate Sensitivities,
    we analysed reef coral cover time series and the ability of reefs to maintain positive carbonate
    production budgets under future conditions. Contrary to expectations, larval connectivity
    patterns did not correlate with increases in maintenance of positive carbonate budgets.
    Instead, reef depth emerged as the primary predictor of cover-time series resilience and
    maintenance of positive carbonate budgets, with deeper reefs benefiting from thermal
    refugia that reduce bleaching impacts. Interaction between reef depth and incoming
    connectivity was found to influence reef temporal behaviour at 15m depth. Our findings suggest that
    depth provides critical protection against thermal stress, with high incoming connectivity enhancing the
    performance of deep reefs, contrasting current intervention
    practices that primarily target easily accessible sites in the 1-10m depth range. These results have
    important implications for reef management, indicating that depth refugia should be
    prioritized for further consideration in conservation planning alongside connectivity
    considerations to maximize coral survival under continuing climate change.
keywords:
  - Coral reef
  - Climate change
  - Ecosystem resilience
  - Reef depth
  - Reef connectivity
date: last-modified
format:
  elsevier-docx:
    journal:
      name: Journal Name
      formatting: preprint
      cite-style: authoryear
bibliography: references.bib
engine: julia
crossref:
  custom:
  - kind: float
    reference-prefix: SFigure
    key: sfig
  - kind: float
    reference-prefix: STable
    key: stbl
editor:
  markdown:
    wrap: 92
    canonical: true
---

@@

**Benjamin Grier:** conceptualisation, data curation, formal analysis, investigation,
methodology, visualization, writing – original draft, writing – review and editing. **Takuya
Iwanaga:** conceptualisation, data curation, formal analysis, investigation, methodology,
visualization, writing – original draft, writing – review and editing. **\[BG: To be
completed.\] ...**

@@

We acknowledge the contributions of Barbara Robson and Jessica Benthuysen to manuscript
preparation. This research was funded by the Reef Restoration and Adaptation Program, and
facilitated by the Australian Institute of Marine Science. We acknowledge thermal projection
data sourced from HighResCoralStress datasets, model parameterisation data sourced from the
Ecological Intelligence for Reef Restoration (EcoRRAP) subprogram, larval transition
probability data derived from the eReefs GBR1 hydrodynamic model, and spatial and depth data
available from the Great Barrier Reef Marine Park Authority. All authors are funded by the
Reef Restoration and Adaptation Program **\[BG: To be completed.\]**.
